Density-dependent selection and the limits of relative fitness

### 2017-12-12 manuscript.tex
l

\textit{Keywords}: $r$/$K$ selection, eco-evolutionary synthesis.

@@

TBA

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of the time-dependent process of selection itself, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.

@@

It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

Having highlighted this redeeming quality of Eq.~\eqref{eq:canonical}, we now turn to its breakdown. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In our lottery model, this can only occur if types differ in $c$ as well as $b$ or $d$. Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}? To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded). The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).

By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent. This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1976}. Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

>>>>

\textit{Keywords}: Lottery model, Wright-Fisher model, $r$/$K$-selection, $\alpha$-selection, eco-evolutionary synthesis.

@@

Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics. Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher. Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present. We then explore the population ecology of relative-fitness using this model as a reference point. Although density-dependent selection causes relative fitness to break down, we emphasize that relative fitness models are fairly robust in populations close to demographic equilbrium. In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent. Our generalized Wright-Fisher model clearly distinguishes these effects, whereas the classical literature on density-dependent selection confounds them. We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models. Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilbrium.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. [will make this sound less dismissive] Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of the time-dependent process of selection itself, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.

@@

It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' [thanks for the suggestion, still need to do references] of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In our lottery model, this can only occur if types differ in $c$ as well as $b$ or $d$. Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}? To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded). The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).

By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent. This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

Note that while this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$, which does not affect $N$.
